ManyTypes4Py: A Benchmark Python Dataset for Machine Learning-based Type Inference\\

### 2021-03-09 paper/main.tex
Minor revision to the MSR'21 paper for publishing

\begin{table*}[!t]
	\centering
	\label{tab:char-dataset}
	\caption{Characteristics of the ManyTypes4Py Dataset}
	\begin{threeparttable}
		\begin{tabular}{l c c c c}
			\toprule
			Metrics & \multicolumn{4}{@{} c @{}}{Dataset}  \\
			\cmidrule{2-5}
			& All & Training & Validation & Test \\
			\midrule
			Repositories\tnote{a} & 5,382 & 4,913 & 2,789 & 3,796 \\
			Lines of code\tnote{b} & 22M & - & - & - \\
			\midrule
			Files & 183,916  & 132,409 & 14,675 & 36,832 \\
			...with type annotations & 50,838 (27.6\%) & 36,542 & 4,105  & 10,191  \\
			\midrule
			Functions & 2,096,797 & 1,509,048 & 169,519 & 418,230 \\
			...with comment & 1,129,573 (53.8\%) & 812,632 & 91,325 & 225,616 \\
			...with return type annotations & 325,532 (15.5\%) & 234,319 & 26,104 & 65,109 \\
			\midrule
			Arguments & 3,923,667 & 2,822,699 & 310,685 & 790,283 \\
			...with comment & 220,976 (5.6\%) & 159,453 & 16,924 & 44,599 \\ 
			...with type annotations & 480,793 (12.2\%) & 347,898 & 37,148 & 95,747 \\
			\midrule
			Types & 869,825 & 347,898 & 89,334 & 192,102 \\
			...unique & 67,060 & 53,614 & 13,995 & 23,572 \\
			\bottomrule
		\end{tabular}
		\begin{tablenotes}
			\item[a] {\footnotesize Note that there is an intersection among repositories in the three sets as the dataset is split by files.}
			\item[b] {\footnotesize Comments and blank lines are ignored when counting lines of code.}
		\end{tablenotes}
	\end{threeparttable}
\end{table*}

@@

\begin{table*}[!t]
	\centering
	\caption{Description of fields in the JSON file of projects produced by the pipeline}
	\label{tab:json-fields}
	\begin{tabular}{c l}
		\toprule
		Field Name in the JSON & Description \\
		\midrule
		\multicolumn{2}{c}{Project}  \\
		\midrule
		\texttt{author/repo} & The name of a project and its author on the GitHub URL  \\
		\midrule
		\texttt{src\_files} & Contains the path of a project's source code files \\
		\midrule
		\texttt{file\_path} & The path of a source code file to differentiate it with other files \\
		\midrule
		\multicolumn{2}{c}{Module}  \\
		\midrule
		\texttt{untyped\_seq} & The normalized seq2seq representation of an analyzed source code file \\
		\midrule
		\texttt{typed\_seq} & Contains the type of identifiers in \texttt{untyped\_seq} if present. Otherwise $0$ is inserted. \\
		\midrule
		\texttt{imports} & Contains the name of imports in an analyzed source code file. \\
		\midrule
		\texttt{variables} & Contains variables' name and their type defined in a module (i.e. global variables) \\
		\midrule
		\texttt{classes} & Contains the JSON object of analyzed classes in a module which is described below \\
		\midrule
		\texttt{funcs} &  Contains the JSON object of analyzed functions in a module, which are described below \\
		\midrule
		\texttt{set} & The set to which a source code file belongs to, i.e., \texttt{train}, \texttt{valid}, \texttt{test} \\
		\midrule
		\multicolumn{2}{c}{Class} \\
		\midrule
		\texttt{name} & The name of an analyzed class in a module \\
		\midrule
		\texttt{variables} & Contains class variables' name and their type if present \\
		\midrule
		\texttt{funcs} & Contains the JSON object of analyzed functions in a class, which are described below \\
		\midrule
		\multicolumn{2}{c}{Function} \\
		\midrule
		\texttt{name} & The name of an analyzed function in either a class or a module \\
		\midrule
		\texttt{params} & Contains an analyzed function's parameter names and their type if present \\
		\midrule
		\texttt{ret\_exprs} & Contains the return expression(s) of an analyzed function \\
		\midrule
		\texttt{ret\_type} & The return type of an analyzed function if present \\
		\midrule
		\texttt{variables} & Contains local variables' name and their type in an analyzed function \\
		\midrule
		\texttt{params\_occur} & Contains parameters and their usages in the body of an analyzed function \\
		\midrule
		\texttt{docstring} & Contains docsting of an analyzed function if present, which has the below subfields \\
		\midrule
		\texttt{docstring.func} & One-line description of an analyzed function if present \\
		\midrule
		\texttt{docstring.ret} & Description of what an analyzed function returns if present \\
		\midrule
		\texttt{docstring.long\_descr} & Long description of an analyzed function if present \\
		\bottomrule
	\end{tabular}
\end{table*}

@@

% Code compeletion
\paragraph{\textbf{Learning-based code completion}} In this application, an ML model is expected to predict part of a word or token for a function or a variable. For DPLs, code completion is a challenging task as there is no type of information available. To overcome this, ASTs are statically analyzed while providing type information \cite{svyatkovskiy2019pythia}. The ManyTypes4Py dataset can be used as a baseline for training a code completion model as it provides partial type annotations for functions and variables. Also, our AST analysis pipeline (LibSA4Py tool) can further be extended to infer types of nodes and variables for simple cases.

@@

\section{Related Work}
There are several Python code corpora that can be used for machine learning-based type inference. Recently, Allamanis et al. \cite{allamanis2020typilus} proposed the Typilus model, which is a graph-based neural model that predicts type annotations for Python. The Typilus model \cite{allamanis2020typilus} is accompanied by a dataset that contains 600 Python projects. Moreover, the source code files of Typilus' dataset are converted to graph representations that are only suitable for training the Typilus model. The ManyTypes4Py dataset provides JSON-formatted analyzed source code files that contains useful type hints for training various machine learning models. Raychev et al. \cite{raychev2016probabilistic} published the Python-150K dataset in 2016, which contains 8,422 Python projects. Unlike our dataset, the Python-150K dataset \cite{raychev2016probabilistic} is not collected solely for the ML-based type inference task, meaning that a large of number projects in the dataset may not have type annotations at all, especially given the time that the dataset was created. Allamanis \cite{allamanis2019adverse} showed that the Python-150K dataset suffers from code duplication despite the removal of project forks.

\section{Conclusion}
We present the ManyTypes4Py dataset, a benchmark Python dataset for ML-based type inference. It contains 5,382 Python projects from GitHub with more than 869K type annotations. The collected Python projects were de-duplicated by removing duplicate source code files to ensure that trained ML models do not have duplication bias. Using the accompanying LibSA4Py tool, the AST of Python source code files were analyzed to provide 16 type hints plus a seq2seq representation for training ML-based type inference models. For each analyzed project, the result of the AST analysis is saved in a JSON-formatted file. Although the dataset's main application is ML-based type inference, it can be a useful baseline for learning-based code completion.

>>>>

\begin{table*}[!t]
	\centering
	\label{tab:char-dataset}
	\caption{Characteristics of the ManyTypes4Py Dataset}
	\begin{threeparttable}
		\begin{tabular}{l l l l l}
			\toprule
			Metrics & \multicolumn{4}{@{} c @{}}{Dataset}  \\
			\cmidrule{2-5}
			& All & Training & Validation & Test \\
			\midrule
			Repositories\tnote{a} & 5,382 & 4,913 & 2,789 & 3,796 \\
			Lines of code\tnote{b} & 22M & - & - & - \\
			\midrule
			Files & 183,916  & 132,409 & 14,675 & 36,832 \\
			...with type annotations & 50,838 (27.6\%) & 36,542 & 4,105  & 10,191  \\
			\midrule
			Functions & 2,096,797 & 1,509,048 & 169,519 & 418,230 \\
			...with comment & 1,129,573 (53.8\%) & 812,632 & 91,325 & 225,616 \\
			...with return type annotations & 325,532 (15.5\%) & 234,319 & 26,104 & 65,109 \\
			\midrule
			Arguments & 3,923,667 & 2,822,699 & 310,685 & 790,283 \\
			...with comment & 220,976 (5.6\%) & 159,453 & 16,924 & 44,599 \\ 
			...with type annotations & 480,793 (12.2\%) & 347,898 & 37,148 & 95,747 \\
			\midrule
			Types & 869,825 & 347,898 & 89,334 & 192,102 \\
			...unique & 67,060 & 53,614 & 13,995 & 23,572 \\
			\bottomrule
		\end{tabular}
		\begin{tablenotes}
			\item[a] {\footnotesize Note that there is an intersection among repositories in the three sets as the dataset is split by files.}
			\item[b] {\footnotesize Comments and blank lines are ignored when counting lines of code.}
		\end{tablenotes}
	\end{threeparttable}
\end{table*}

@@

\begin{table*}[!t]
	\centering
	\caption{Description of fields in the JSON file of projects produced by the LibSA4Py pipeline}
	\label{tab:json-fields}
	\begin{tabular}{c l}
		\toprule
		Field Name in the JSON & Description \\
		\midrule
		\multicolumn{2}{c}{Project}  \\
		\midrule
		\texttt{author/repo} & The name of a project and its author on the GitHub URL  \\
		\midrule
		\texttt{src\_files} & Contains the path of a project's source code files \\
		\midrule
		\texttt{file\_path} & The path of a source code file to differentiate it with other files \\
		\midrule
		\multicolumn{2}{c}{Module}  \\
		\midrule
		\texttt{untyped\_seq} & The normalized seq2seq representation of an analyzed source code file \\
		\midrule
		\texttt{typed\_seq} & Contains the type of identifiers in \texttt{untyped\_seq} if present. Otherwise $0$ is inserted. \\
		\midrule
		\texttt{imports} & Contains the name of imports in an analyzed source code file \\
		\midrule
		\texttt{variables} & Contains variables' name and their type defined in a module (i.e. global variables) \\
		\midrule
		\texttt{classes} & Contains the JSON object of analyzed classes in a module which is described below \\
		\midrule
		\texttt{funcs} &  Contains the JSON object of analyzed functions in a module, which are described below \\
		\midrule
		\texttt{set} & The set to which a source code file belongs to, i.e., \texttt{train}, \texttt{valid}, \texttt{test} \\
		\midrule
		\multicolumn{2}{c}{Class} \\
		\midrule
		\texttt{name} & The name of an analyzed class in a module \\
		\midrule
		\texttt{variables} & Contains class variables' name and their type if present \\
		\midrule
		\texttt{funcs} & Contains the JSON object of analyzed functions in a class, which are described below \\
		\midrule
		\multicolumn{2}{c}{Function} \\
		\midrule
		\texttt{name} & The name of an analyzed function in either a class or a module \\
		\midrule
		\texttt{params} & Contains an analyzed function's parameter names and their type if present \\
		\midrule
		\texttt{ret\_exprs} & Contains the return expression(s) of an analyzed function \\
		\midrule
		\texttt{ret\_type} & The return type of an analyzed function if present \\
		\midrule
		\texttt{variables} & Contains local variables' name and their type in an analyzed function \\
		\midrule
		\texttt{params\_occur} & Contains parameters and their usages in the body of an analyzed function \\
		\midrule
		\texttt{docstring} & Contains docsting of an analyzed function if present, which has the below subfields \\
		\midrule
		\texttt{docstring.func} & One-line description of an analyzed function if present \\
		\midrule
		\texttt{docstring.ret} & Description of what an analyzed function returns if present \\
		\midrule
		\texttt{docstring.long\_descr} & Long description of an analyzed function if present \\
		\bottomrule
	\end{tabular}
\end{table*}

@@

% Code compeletion
\paragraph{\textbf{Learning-based code completion}} In this application, an ML model is expected to predict part of a word or token for a function or a variable. For DPLs, code completion is a challenging task as there is no type information available. To overcome this, ASTs are statically analyzed while providing type information \cite{svyatkovskiy2019pythia}. The ManyTypes4Py dataset can be used as a baseline for training a code completion model as it provides partial type annotations for functions and variables. Also, our AST analysis pipeline (LibSA4Py tool) can further be extended to infer types of nodes and variables for simple cases.

@@

\section{Related Work}
There are several Python code corpora that can be used for machine learning-based type inference. Recently, Allamanis et al. \cite{allamanis2020typilus} proposed the Typilus model, which is a graph-based neural model that predicts type annotations for Python. The Typilus model \cite{allamanis2020typilus} is accompanied by a dataset that contains 600 Python projects. Moreover, the source code files of Typilus' dataset are converted to graph representations that are only suitable for training the Typilus model. The ManyTypes4Py dataset provides JSON-formatted analyzed source code files that contains useful type hints for training various machine learning models. Raychev et al. \cite{raychev2016probabilistic} published the Python-150K dataset in 2016, which contains 8,422 Python projects. Unlike our dataset, the Python-150K dataset \cite{raychev2016probabilistic} is not collected solely for the ML-based type inference task, meaning that a large number of projects in the dataset may not have type annotations at all, especially given the time that the dataset was created. Allamanis \cite{allamanis2019adverse} showed that the Python-150K dataset suffers from code duplication despite the removal of project forks.

\section{Conclusion}
In this paper, we present the ManyTypes4Py dataset, a benchmark Python dataset for ML-based type inference. It contains 5,382 Python projects from GitHub with more than 869K type annotations. The collected Python projects were de-duplicated by removing duplicate source code files to ensure that trained ML models do not have duplication bias. Using the accompanying LibSA4Py tool, the AST of Python source code files were analyzed to provide 16 type hints plus a seq2seq representation for training ML-based type inference models. For each analyzed project, the result of the AST analysis is saved in a JSON-formatted file. Although the dataset's main application is ML-based type inference, it can be a useful baseline for learning-based code completion.
